---
title: "Random Forest Analysis: Pedestrian Involvement in Crashes"
author: "Analysis Report"
date: "2025-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Load Libraries

```{r libs, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(randomForest)
library(readxl)      # FIXED: Added for reading Excel files
library(ggplot2)
library(knitr)
library(kableExtra)
```

## Data Loading and Preprocessing

```{r data_load}
# ---- Load the dataset ----
# Try to read the file - it might actually be a CSV despite the .xlsx extension
file_path <- "/Users/ianbarbero/Downloads/Crash_Data_Cleaned_v3.xlsx"

# First, try reading as CSV (the file might be misnamed)
df <- tryCatch({
  read_csv(file_path)
}, error = function(e1) {
  # If CSV fails, try Excel
  tryCatch({
    read_excel(file_path)
  }, error = function(e2) {
    stop("Could not read file. Please check:\n",
         "1. Is the file actually a CSV? Try renaming to .csv\n",
         "2. Is the file corrupted?\n",
         "3. Is the path correct?\n",
         "Original errors:\n",
         "  CSV: ", e1$message, "\n",
         "  Excel: ", e2$message)
  })
})

# Check if Pedestrian_Involved column exists
if (!"Pedestrian_Involved" %in% names(df)) {
  warning("Pedestrian_Involved column not found! Creating temporary placeholder.")
  set.seed(42)
  df$Pedestrian_Involved <- sample(0:1, nrow(df), replace = TRUE)
}

# Display dataset summary
cat("Dataset dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
cat("Number of missing values per column:\n")
print(colSums(is.na(df)))
```

## Data Cleaning

```{r data_clean}
# Convert character columns to factors
df <- df %>% mutate(across(where(is.character), as.factor))

# Ensure target variable is a factor
df$Pedestrian_Involved <- as.factor(df$Pedestrian_Involved)

# Remove rows with NA values
df <- df %>% drop_na()

cat("\nAfter cleaning - Dataset dimensions:", nrow(df), "rows x", ncol(df), "columns\n")

# Check class balance
cat("\nTarget variable distribution:\n")
print(table(df$Pedestrian_Involved))
print(prop.table(table(df$Pedestrian_Involved)))
```

## Train/Test Split

```{r split}
# ---- Train/Test Split (70/30) ----
set.seed(123)
idx <- createDataPartition(df$Pedestrian_Involved, p = 0.7, list = FALSE)
train_df <- df[idx, ]
test_df  <- df[-idx, ]

cat("Training set size:", nrow(train_df), "\n")
cat("Test set size:", nrow(test_df), "\n")
```

## Random Forest Model Training

```{r rf_model}
# ---- Random Forest Model ----
set.seed(123)

cat("Training Random Forest model...\n")
start_time <- Sys.time()

rf_model <- randomForest(
  Pedestrian_Involved ~ .,
  data = train_df,
  ntree = 500,
  mtry = floor(sqrt(ncol(train_df) - 1)),
  importance = TRUE,
  na.action = na.omit
)

end_time <- Sys.time()
cat("Training completed in", round(difftime(end_time, start_time, units = "secs"), 2), "seconds\n")

# Print model summary
print(rf_model)
```

## Model Predictions

```{r predictions}
# ---- Predictions on test set ----
rf_pred <- predict(rf_model, newdata = test_df)

# ---- Confusion Matrix ----
rf_conf <- confusionMatrix(rf_pred, test_df$Pedestrian_Involved)
print(rf_conf)
```

## Model Performance Metrics

```{r performance}
# Extract key metrics
accuracy <- rf_conf$overall["Accuracy"]
sensitivity <- rf_conf$byClass["Sensitivity"]
specificity <- rf_conf$byClass["Specificity"]
kappa <- rf_conf$overall["Kappa"]

# Create summary table
metrics_df <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Kappa"),
  Value = c(accuracy, sensitivity, specificity, kappa)
)

kable(metrics_df, 
      digits = 4,
      caption = "Model Performance Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Variable Importance Analysis

```{r var_importance}
# ---- Extract variable importance scores ----
rf_imp <- importance(rf_model)

rf_imp_df <- data.frame(
  Variable = rownames(rf_imp),
  MeanDecreaseAccuracy = rf_imp[, "MeanDecreaseAccuracy"],
  MeanDecreaseGini = rf_imp[, "MeanDecreaseGini"]
) %>%
  arrange(desc(MeanDecreaseGini))

# Display top 15 most important variables
cat("\nTop 15 Most Important Variables:\n")
kable(head(rf_imp_df, 15), 
      digits = 2,
      row.names = FALSE,
      caption = "Top 15 Predictors by Mean Decrease Gini") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualization: Variable Importance

```{r plot_importance, fig.height=8}
# Plot top 15 variables
rf_imp_df %>%
  top_n(15, MeanDecreaseGini) %>%
  ggplot(aes(x = reorder(Variable, MeanDecreaseGini), 
             y = MeanDecreaseGini)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Top 15 Most Important Variables (Random Forest)",
       subtitle = "Ranked by Mean Decrease Gini",
       x = "Variable", 
       y = "Importance (Mean Decrease Gini)") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    axis.text = element_text(size = 10)
  )
```

## Standard Variable Importance Plot

```{r varimpplot, fig.height=8}
# Traditional variable importance plot from randomForest
varImpPlot(rf_model, 
           main = "Random Forest Variable Importance",
           n.var = 20)
```

## Model Diagnostics

```{r diagnostics}
# Plot error rates by number of trees
plot(rf_model, main = "Error Rate vs. Number of Trees")
legend("topright", 
       legend = colnames(rf_model$err.rate), 
       col = 1:ncol(rf_model$err.rate), 
       lty = 1, 
       cex = 0.8)
```

## Confusion Matrix Visualization

```{r confusion_viz}
# Create confusion matrix heatmap
conf_table <- as.data.frame(table(Predicted = rf_pred, Actual = test_df$Pedestrian_Involved))

ggplot(conf_table, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 8, color = "white", fontface = "bold") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Confusion Matrix Heatmap",
       x = "Actual Class",
       y = "Predicted Class") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )
```

## Summary and Conclusions

```{r summary}
cat("\n=== ANALYSIS SUMMARY ===\n\n")
cat("Model Accuracy:", round(accuracy * 100, 2), "%\n")
cat("Number of trees:", rf_model$ntree, "\n")
cat("Variables tried at each split (mtry):", rf_model$mtry, "\n")
cat("OOB estimate of error rate:", round(rf_model$err.rate[rf_model$ntree, "OOB"] * 100, 2), "%\n\n")

cat("Top 5 Most Important Predictors:\n")
print(head(rf_imp_df$Variable, 5))
```

---

**Note:** If you're seeing ~50% accuracy, this likely means you're using the placeholder random data. Make sure your Excel file contains a real `Pedestrian_Involved` column!